Is There a Competitive Advantage to Using Multivariate Statistical or Machine Learning Methods Over the Bross Formula in the hdPS Framework for Bias and Variance Estimation?

### 2024-09-14 Writeup/BMC admin docs/tex/manuscript.tex
bmc submission

\abstract{\textbf{Purpose}: We aims to evaluate various proxy selection
methods within the context of high-dimensional propensity score (hdPS)
analysis. The study focuses on assessing the performance of these
methods, including alternative variable selection approaches, in
selecting proxy variables for confounding adjustment compared to the
traditional hdPS method that is rooted in Bross formula. The goal is to
understand better the performance of these alternative methods in
estimating treatment effects, and identify scenarios in which they may
perform better. \textbf{Methods:} Using data from three cycles of the
National Health and Nutrition Examination Survey (NHANES) spanning
2013-2018, we motivated the study by examining the association between
obesity and diabetes. A plasmode simulation framework based on this data
was employed to mimic real-world data structures. Simulations were
conducted under three scenarios: Frequent Exposure and Outcome, Rare
Exposure and Frequent Outcome, and Frequent Exposure and Rare Outcome.
The performance of a variety of proxy selection methods---including
tree-based methods, LASSO-based methods, and the Genetic Algorithm
(GA)---was evaluated across these scenarios using standard simulation
metrics. \textbf{Results:} XGBoost consistently demonstrated the lowest
MSE and high coverage, making it a reliable method overall, although it
did not always exhibit the lowest bias. In contrast, GA consistently
showed the highest bias and MSE, with lower coverage and greater
variability, indicating its unsuitability for accurate effect
estimation. The kitchen sink model, Bross-based hdPS, and Hybrid hdPS
methods performed moderately well, with low bias and moderate MSE, but
varied in coverage across scenarios. Scenario-specific trends revealed
that rare outcome scenarios yielded lower MSE and better precision,
while rare exposure scenarios were associated with higher bias and MSE.
\textbf{Conclusion:} Aside from GA, the performance of most other
methods in your study appears to be comparable in each scenario, though
with some variations depending on specific metrics. The findings
underscore the importance of selecting appropriate methods for hdPS
analysis based on the characteristics of the data, particularly the
prevalence of exposure and outcome. While XGBoost excelled in overall
accuracy and precision, the choice of method should be tailored to the
specific epidemiological goal to optimize bias, coverage, and precision
in estimating treatment effects.}

@@

\textbf{High-dimensional Propensity Score (hdPS) Algorithm}: In
epidemiology, proxy variables are commonly used as substitutes for
confounders that are difficult or impossible to measure directly, such
as socioeconomic status, lifestyle factors, or health behaviors
\citep{vanderweele2019principles}. The high-dimensional propensity score
(hdPS) is a pharmacoepidemiological method designed to reduce
confounding bias in large healthcare databases
\citep{schneeweiss2009high}. Unlike traditional propensity score models
that rely on investigator-specified or manually selected covariates,
hdPS automatically ranks a wide array of proxy variables from healthcare
records---such as diagnosis codes, medications, and procedures---using
the Bross formula \citep{wyss2018erratum, bross1966spurious}. The Bross
formula ranks these variables based on their marginal associations with
both exposure and outcome. These selected proxy variables serve as
surrogates for unmeasured or poorly measured confounders, helping reduce
bias in treatment effect estimates. The hdPS algorithm further refines
the selection by prioritizing variables based on their prevalence and
potential for confounding, defined by their association with both
exposure and outcome \citep{schneeweiss2009high}.

\textbf{Multivariate Machine Learning Extensions}: Although the Bross
formula performs well in certain contexts, it has limitations in
capturing complex interactions, nonlinearities, and higher-order
associations between variables, especially in high-dimensional settings
where it does not account for the multivariate structure of other
covariates \citep{karim2024high, karim2018can}. To address these
limitations, multivariate machine learning methods such as LASSO,
Elastic Net, and Random Forests have been applied within the hdPS
framework. These methods are better suited for high-dimensional data,
where they can more effectively handle complex relationships and improve
the selection of proxy variables, thus enhancing the precision of
treatment effect estimates
\citep{karim2018can, schneeweiss2017variable, franklin2015regularized}.
Simulation studies and empirical research have shown that these machine
learning-based methods, or hybrid approaches combining the Bross formula
with machine learning, can reduce confounding more effectively and
increase efficiency compared to the Bross formula alone in certain
settings
\citep{karim2018can, schneeweiss2017variable, franklin2015regularized}.

\textbf{Assessing the Simulation Performance}: Previous studies have
often focused on Mean Squared Error (MSE) as a primary metric for
evaluating the performance of hdPS and its machine learning extensions
\citep{karim2018can, franklin2015regularized}. However, in
high-dimensional settings with singly robust methods, such as hdPS and
machine learning approaches like LASSO, MSE may not always be the most
reliable measure. MSE combines both bias and variance into a single
metric, which makes interpretation challenging when variance estimation
is unstable---a common issue in these methods. In contrast, coverage,
which measures the proportion of confidence intervals that capture the
true treatment effect, provides a more direct and meaningful assessment
of a model's reliability. In realistic observational studies, where
model misspecification is often inevitable, coverage---along with
related metrics such as bias-eliminated coverage and relative error in
standard error (SE; which compares model-based SE with empirical
SE)---can reveal whether confidence intervals or SEs are too narrow
(underestimating uncertainty) or too wide (overestimating uncertainty).
This insight is crucial in determining whether the model delivers valid
estimates despite misspecification. Even a model with poor MSE but good
coverage may still be valuable, as it produces realistic confidence
intervals. By shifting the focus to coverage, rather than relying solely
on MSE, we can achieve a more comprehensive understanding of method
performance, especially in cases where unstable variance estimation
might distort conclusions drawn from MSE alone.

@@

\textbf{Plasmode simulation}: To rigorously assess the performance of
the methods under consideration, we employed a plasmode simulation
framework, which is particularly well-suited for reflecting real-world
data structures and complexities \citep{franklin2014plasmode}. This
approach was modeled after the analytic dataset derived from NHANES and
involved resampling from the observed covariates and exposure
information (i.e., obesity) without altering them. By mirroring key
aspects of an actual epidemiological study, this simulation framework
offers a substantial advantage over traditional Monte Carlo simulations,
which often rely on hypothetical assumptions.

@@

\textbf{Performance Measures}: From this simulation, we derived several
performance metrics to evaluate the effectiveness of the methods under
consideration: (1) bias, (2) average model standard error (SE; the
average of estimated SEs obtained from a model over repeated samples),
(3) empirical SE (the standard deviation of estimated treatment effects
across repeated samples), (4) mean squared error (MSE), (5) coverage
probability of 95\% confidence intervals, (6) bias-corrected coverage,
and (7) Zip plot \citep{morris2019using, white2023check}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Kitchen sink model}: This is a base model for comparison,
  where no variable selection approaches were used. All
  investigator-selected features and all proxy variables were used to
  model \citep{karim2018can}.
\item
  \textbf{hdPS using Bross formula}: The Bross formula is a statistical
  method used to calculate the bias introduced by not adjusting for a
  covariate \citep{bross1966spurious}. In hdPS analysis, this formula
  was originally applied to each proxy variable to measure and rank the
  potential bias if the covariate were not adjusted for. In our
  analysis, the 100 proxies with the highest bias rankings are selected
  for further modeling \citep{schneeweiss2009high, wyss2018erratum}.
\item
  \textbf{Least Absolute Shrinkage and Selection Operator (LASSO)}:
  LASSO is a variable selection technique that limits the number of
  variables by adding a penalty term to the regression model.
  Cross-validation (CV) is used in LASSO to identify variables with
  non-zero coefficients in the best model by optimizing the penalty
  value
  \citep{franklin2015regularized, schneeweiss2017variable, karim2018can}.
\item
  \textbf{Hybrid of hdPS and LASSO}: Instead of relying solely on LASSO
  for variable selection, a hybrid approach combines the Bross formula
  and LASSO. First, hdPS variables are selected using the hdPS algorithm
  (e.g., the top 100), and then LASSO is applied to further refine the
  selection \citep{karim2018can, franklin2015regularized}.
\item
  \textbf{Elastic Net}: Elastic Net is an extension of LASSO that
  includes an additional penalty term to handle multicollinearity by
  grouping correlated features and selecting the most representative
  ones \citep{karim2018can}.
\item
  \textbf{Random Forest}: The Random Forest (RF) algorithm is an
  ensemble learning method that constructs multiple decision trees to
  perform classification \citep{breiman2001random}. It calculates the
  importance of each proxy variable based on the decrease in impurity or
  Gini importance, providing a ranking of the proxies. The top 100
  variables from this ranking are manually selected for further modeling
  \citep{schneeweiss2017variable}.
\item
  \textbf{XGBoost}: XGBoost is a gradient boosting algorithm used to
  optimize machine learning models \citep{chen2016xgboost}. It builds
  decision trees that make splits based on maximum impurity reduction,
  and it assigns an importance score to each proxy variable by
  calculating the mean decrease in impurity
  \citep{xiao2024interpretable}.
\item
  \textbf{Stepwise}: Stepwise selection is a progressive feature
  selection method that can proceed in two directions---forward or
  backward---based on the maximum adjusted R-squared. We have
  implemented two versions: (a) Forward selection (FS) starts with an
  initial model (e.g., including all investigator-selected features) and
  adds proxies to the model one at a time. (b) Backward elimination (BE)
  starts with a full model (e.g., all investigator-selected features and
  all proxy variables) and removes features one at a time based on their
  contribution to the model.
\item
  \textbf{Genetic algorithm (GA)}: GA is an evolutionary algorithm
  inspired by the theory of natural selection
  \citep{holland1975adaptation}. It operates by evolving offspring from
  a population of the fittest individuals over several generations,
  evaluating and selecting the best combination of features or variables
  that maximize prediction accuracy.
\end{enumerate}

@@

The results for each method under the different scenarios are summarized
below. See Figures \ref{fig:bias-comparison} and
\ref{fig:Coverage-comparison} for an overview of the performance in
terms of bias and coverage, respectively.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: Bross-based hdPS exhibited the smallest bias (-0.0001),
  and the kitchen sink model (0.0002) was the second. GA shows the
  highest bias (0.0287), indicating a substantial deviation from the
  true effect. Among the other methods, Hybrid hdPS (0.0016), and
  Elastic Net (0.0036) demonstrated low bias. XGBoost (0.0074) had
  slightly higher bias than Random Forest (RF) (0.0034).
\item
  \emph{Coverage}: The coverage for most methods was high, with Hybrid
  hdPS, Forward Selection, Backward Elimination, LASSO, and Elastic Net
  achieving values around 98\%, indicating well-calibrated confidence
  intervals. However, GA had noticeably lower coverage (83.8\%),
  indicating that its confidence intervals might be too narrow or
  biased, potentially missing the true effect. After applying bias
  elimination (as there were substantial bias associated with this
  method), GA's coverage improved to 96\%.
\item
  \emph{Mean Squared Error (MSE)}: XGBoost achieved the lowest MSE
  (0.0006). GA maintained the highest MSE (0.0016), reflecting its
  higher bias and variability. The kitchen sink model (0.0009),
  Bross-based hdPS (0.0008), Hybrid hdPS (0.0008), and Elastic Net
  (0.0009) all had relatively similar and moderate MSE values.
\item
  \emph{Standard Error (SE)}: XGBoost exhibited the lowest Empirical SE
  (0.0229), indicating high precision in its estimates. The kitchen sink
  model had the highest Empirical SE (0.0305), suggesting greater
  variability. Other methods, including GA (0.0274), Hybrid hdPS
  (0.0278), and Bross-based hdPS (0.0287), showed moderate variability.
  LASSO (0.0299) and Elastic Net (0.0294) had slightly higher
  variability. The Model-based SE followed a similar pattern, with
  XGBoost (0.0268) showing the lowest variability and the kitchen sink
  model (0.0333) the highest, indicating less precision in its
  estimates. When comparing relative error in SE estimation, XGBoost
  performed the worst. See Appendix \S C for further details.
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: The kitchen sink model showed a relatively low bias
  (0.0025), while GA continued to exhibit the highest bias (0.0408),
  indicating a significant deviation from the true effect. XGBoost had a
  bias of 0.0259, which, while still higher than other methods, but was
  lower than GA. Other methods, such as Bross-based hdPS (0.0035),
  Hybrid hdPS (0.0049), and Elastic Net (0.0053), demonstrated moderate
  bias. Random Forest (RF) (0.0127) and Forward Selection (0.0108) had
  slightly higher bias but remained within an acceptable range.
\item
  \emph{Coverage}: Coverage levels remained high for most methods, with
  XGBoost achieving the highest coverage at 96.2\%, indicating
  well-calibrated confidence intervals despite its higher bias. The GA
  method had lower coverage (92.2\%), suggesting that its confidence
  intervals might be narrower, potentially missing the true effect.
  Other methods such as RF, Forward Selection, Backward Elimination, and
  Hybrid hdPS maintained coverage values around 94-95\%, suggesting
  adequate interval calibration. Bias-eliminated coverage for GA
  improved to 94.2\%, but it was still slightly lower than other methods
  (e.g., forward selection).
\item
  \emph{Mean Squared Error (MSE)}: Forward selection demonstrated the
  lowest MSE (0.0030), and then Hybrid hdPS and XGBoost. The GA method
  had a higher MSE (0.0043), reflecting its substantial bias and
  variability. The kitchen sink model also had an MSE of 0.0043, similar
  to GA, while other methods like Bross-based hdPS (0.0035), RF
  (0.0039), and Elastic Net (0.0036) exhibited moderate MSE values,
  indicating reasonable accuracy.
\item
  \emph{Standard Error (SE)}: The lowest Empirical SE was observed with
  XGBoost (0.0507) and GA (0.0510), reflecting high precision despite
  their higher bias. The kitchen sink model had the highest Empirical SE
  (0.0656), indicating greater variability. Hybrid hdPS (0.0564),
  Bross-based hdPS (0.0595), and RF (0.0609) showed moderate
  variability. Forward Selection (0.0537) and Backward Elimination
  (0.0576) had lower variability compared to the kitchen sink model. In
  terms of Model-based SE, XGBoost (0.0531) and GA (0.0533) continued to
  show low variability, while the kitchen sink model had the highest
  Model-based SE (0.0623), indicating less precision in its estimates.
  When comparing relative error in SE estimation, XGBoost and kitchen
  sink model performed the worst (in other direction).
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: In this scenario, the kitchen sink model exhibited a
  moderate negative bias (-0.0093), similar to the Bross-based hdPS
  method (-0.0088). GA showed a significantly higher bias (0.0362),
  indicating a substantial deviation from the true effect. Among other
  methods, XGBoost demonstrated the lowest bias (-0.0061), while methods
  like Hybrid hdPS (-0.0082), Forward Selection (-0.0070), and Backward
  Elimination (-0.0070) had slightly higher but still moderate biases.
  Elastic Net and LASSO both had biases of -0.0079, reflecting slightly
  larger deviations compared to XGBoost but still within acceptable
  limits.
\item
  \emph{Coverage}: Most methods achieved good coverage, with XGBoost,
  RF, and Forward Selection each achieving a coverage rate of 95.4\%,
  indicating well-calibrated confidence intervals. The GA method,
  however, had slightly lower coverage (91.8\%), indicating that its
  confidence intervals might be narrower, potentially excluding the true
  effect. Bross-based hdPS and the kitchen sink model had slightly lower
  coverage values of 93.8\% and 93.4\%, respectively. After accounting
  for bias, the bias-eliminated coverage for most methods, except GA,
  remained high, with values ranging from 98.4\% to 99.0\%, indicating
  that most methods effectively adjusted for bias in their coverage
  estimates. GA's bias-eliminated coverage was lower at 93.4\%,
  reflecting its higher inherent bias.
\item
  \emph{Mean Squared Error (MSE)}: XGBoost exhibited the lowest MSE
  (0.0003). GA had the highest MSE (0.0040), reflecting its substantial
  bias and variability. The kitchen sink model (0.0005), Bross-based
  hdPS (0.0005), and other methods like Hybrid hdPS (0.0004) and Elastic
  Net (0.0005) all had relatively similar MSE values, indicating
  moderate accuracy.
\item
  \emph{Standard Error (SE)}: The lowest Empirical SE was observed with
  XGBoost (0.0152), reflecting high precision in its estimates. The GA
  method exhibited the highest Empirical SE (0.0523), indicating greater
  variability and less precision. Methods such as Hybrid hdPS (0.0184),
  Forward Selection (0.0187), and Elastic Net (0.0203) showed moderate
  variability, while Bross-based hdPS (0.0206) and the kitchen sink
  model (0.0212) had slightly higher variability. In terms of
  Model-based SE, XGBoost (0.0179) again showed the lowest variability,
  consistent with its low Empirical SE, indicating that it provided the
  most stable estimates. The kitchen sink model had a slightly higher
  Model-based SE (0.0219), indicating less precision in its estimates.
  When comparing relative error in SE estimation, XGBoost performed the
  worst.
\end{enumerate}

@@

@@

See Figure \ref{fig:comparison-plot-updated} for the results from
analyzing the NHANES (2013-2018) dataset. The methods are arranged
according to the number of selected proxy variables. Among all variable
selection algorithms, Random Forest (RF) and XGBoost demonstrate the
highest odds ratios (ORs), with values of 1.59 and 1.56, respectively.
The ORs for the remaining methods cluster around 1.52. Additionally,
with the exception of RF and the Bross formula hdPS, a general pattern
emerges where methods selecting a larger number of proxy variables yield
lower ORs. For risk difference (RD), RF and XGBoost also exhibit the
highest value 0.08, while the remaining methods converge around 0.077.
The trend observed in the OR results appears to persist in the RD
analysis.

@@

{\centering \includegraphics[width=1\linewidth,]{manuscript_files/figure-latex/unnamed-chunk-3-1}

>>>>

\abstract{\textbf{Purpose}: We aim to evaluate various proxy selection
methods within the context of high-dimensional propensity score (hdPS)
analysis. This study aimed to systematically evaluate and compare the
performance of traditional statistical methods and machine learning
approaches within the hdPS framework, focusing on key metrics such as
bias, standard error (SE), and coverage, under various exposure and
outcome prevalence scenarios. \textbf{Methods:} We conducted a plasmode
simulation study using data from the National Health and Nutrition
Examination Survey (NHANES) cycles from 2013 to 2018. We compared
methods including the kitchen sink model, Bross-based hdPS, Hybrid hdPS,
LASSO, Elastic Net, Random Forest, XGBoost, and Genetic Algorithm (GA).
The performance of each method was assessed based on bias, MSE, coverage
probability, and SE estimation across three epidemiological scenarios:
frequent exposure and outcome, rare exposure and frequent outcome, and
frequent exposure and rare outcome. \textbf{Results:} XGBoost
consistently demonstrated strong performance in terms of MSE and
coverage, making it effective for scenarios prioritizing precision.
However, it exhibited higher bias, particularly in rare exposure
scenarios, suggesting it is less suited when minimizing bias is
critical. In contrast, GA showed significant limitations, with
consistently high bias and MSE, making it the least reliable method.
Bross-based hdPS, and Hybrid hdPS methods provided a
balanced approach, with low bias and moderate MSE, though coverage
varied depending on the scenario. Rare outcome scenarios generally
resulted in lower MSE and better precision, while rare exposure
scenarios were associated with higher bias and MSE. Notably, traditional
statistical approaches such as forward selection and backward
elimination performed comparably to more sophisticated machine learning
methods in terms of bias and coverage, suggesting that these simpler
approaches may be viable alternatives due to their computational
efficiency. \textbf{Conclusion:} The results highlight the importance of
selecting hdPS methods based on the specific characteristics of the
data, such as exposure and outcome prevalence. While advanced machine
learning methods such as XGBoost can enhance precision, simpler methods
such as forward selection or backward elimination may offer similar
performance in terms of bias and coverage with fewer computational
demands. Tailoring the choice of method to the epidemiological scenario
is essential for optimizing the balance between bias reduction and
precision.}

@@

\textbf{High-dimensional Propensity Score (hdPS) Algorithm}: In
epidemiology, proxy variables are commonly used as substitutes for
confounders that are difficult or impossible to measure directly, such
as socioeconomic status, lifestyle factors, or health behaviors
\citep{vanderweele2019principles}. The hdPS is a pharmacoepidemiological
method designed to reduce confounding bias in large healthcare databases
\citep{schneeweiss2009high}. Unlike traditional propensity score models
that rely on investigator-specified or manually selected covariates,
hdPS automatically ranks a wide array of proxy variables from healthcare
records---such as diagnosis codes, medications, and procedures---using
the Bross formula \citep{wyss2018erratum, bross1966spurious}. The Bross
formula ranks these variables based on their marginal associations with
both exposure and outcome. These selected proxy variables serve as
surrogates for unmeasured or poorly measured confounders, helping reduce
bias in treatment effect estimates. The hdPS algorithm further refines
the selection by prioritizing variables based on their prevalence and
potential for confounding, defined by their association with both
exposure and outcome \citep{schneeweiss2009high}.

\textbf{Multivariate Machine Learning Extensions}: Although the Bross
formula performs well in certain contexts, it has limitations in
capturing complex interactions, nonlinearities, and higher-order
associations between variables, especially in high-dimensional settings
where it does not account for the multivariate structure of other
covariates \citep{karim2024high, karim2018can}. To address these
model-specification-related limitations, multivariate machine learning
methods such as LASSO, Elastic Net, and Random Forests have been applied
within the hdPS framework. These methods are better suited for
high-dimensional data, where they can more effectively handle complex
relationships and improve the selection of proxy variables, thus
enhancing the precision of treatment effect estimates
\citep{karim2018can, schneeweiss2017variable, franklin2015regularized}.
Simulation studies and empirical research have shown that these machine
learning-based methods, or hybrid approaches combining the Bross formula
with machine learning, can reduce confounding more effectively and
increase efficiency compared to the Bross formula alone in certain
settings
\citep{karim2018can, schneeweiss2017variable, franklin2015regularized}.

\textbf{Assessing the Simulation Performance}: Other than bias, previous
studies have primarily focused on Mean Squared Error (MSE) as a key
metric for evaluating the performance of hdPS and its machine learning
extensions
\citep{franklin2015regularized, pang2016effect, wyss2018using, karim2018can, simon2023evaluating}.
However, in high-dimensional settings with singly robust methods, such
as hdPS and machine learning approaches such as LASSO, MSE may not
always be the most reliable measure. MSE combines both bias and variance
into a single metric, which makes interpretation challenging when
variance estimation is unstable---a common issue in these methods. In
contrast, coverage, which measures the proportion of confidence
intervals that capture the true treatment effect, provides a more direct
and meaningful assessment of a model's reliability. In realistic
observational studies, where model misspecification is often inevitable,
coverage---along with related metrics such as bias-eliminated coverage
and relative error in standard error (SE; which compares model-based SE
with empirical SE)---can reveal whether confidence intervals or SEs are
too narrow (underestimating uncertainty) or too wide (overestimating
uncertainty). This insight is crucial in determining whether the model
delivers valid estimates despite misspecification. Even a model with
poor MSE but good coverage may still be valuable, as it produces
realistic confidence intervals. By shifting the focus to coverage,
rather than relying solely on MSE, we can achieve a more comprehensive
understanding of method performance, especially in cases where unstable
variance estimation might distort conclusions drawn from MSE alone.

@@

\textbf{Plasmode simulation}: To rigorously assess the performance of
the methods under consideration, we employed a plasmode simulation
framework, which is particularly well-suited for reflecting real-world
data structures and complexities \citep{franklin2014plasmode}. This
approach was inspired by the analytic dataset derived from NHANES and
involved resampling from the observed covariates and exposure
information (i.e., obesity) without altering them. By mirroring key
aspects of an actual epidemiological study, this simulation framework
offers a substantial advantage over traditional Monte Carlo simulations,
which often rely on hypothetical assumptions.

@@

\textbf{Performance Measures}: From this simulation, we derived several
performance metrics to evaluate the effectiveness of the methods under
consideration: (1) bias, (2) average model-based SE (the average of
estimated SEs obtained from a model over repeated samples), (3)
empirical SE (the standard deviation of estimated treatment effects
across repeated samples), (4) MSE, (5) coverage probability of 95\%
confidence intervals, (6) bias-corrected coverage, and (7) Zip plot
\citep{morris2019using, white2023check}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Kitchen sink model}: This is a base model for comparison,
  where no variable selection approaches were used. All
  investigator-selected features and all proxy variables were used to
  model \citep{karim2018can}.
\item
  \textbf{hdPS using Bross formula}: The Bross formula is a statistical
  method used to calculate the bias introduced by not adjusting for a
  covariate \citep{bross1966spurious}. In hdPS analysis, this formula
  was originally applied to each proxy variable to measure and rank the
  potential bias if the covariate were not adjusted for. In our
  analysis, the 100 proxies with the highest bias rankings are selected
  for further modeling \citep{schneeweiss2009high, wyss2018erratum}.
\item
  \textbf{Least Absolute Shrinkage and Selection Operator (LASSO)}:
  LASSO is a variable selection technique that limits the number of
  variables by adding a penalty term to the regression model.
  Cross-validation (CV) is used in LASSO to identify variables with
  non-zero coefficients in the best model by optimizing the penalty
  value
  \citep{franklin2015regularized, schneeweiss2017variable, karim2018can}.
\item
  \textbf{Hybrid of hdPS and LASSO}: Instead of relying solely on LASSO
  for variable selection, a hybrid approach combines the Bross formula
  and LASSO. First, proxy variables are selected using the hdPS
  algorithm (e.g., the top 100), and then LASSO is applied to further
  refine the selection \citep{karim2018can, franklin2015regularized}.
\item
  \textbf{Elastic Net}: Elastic Net is an extension of LASSO that
  includes an additional penalty term to handle multicollinearity by
  grouping correlated features and selecting the most representative
  ones \citep{karim2018can}.
\item
  \textbf{Random Forest}: The Random Forest algorithm is an ensemble
  learning method that constructs multiple decision trees to perform
  classification \citep{breiman2001random}. It calculates the importance
  of each proxy variable based on the decrease in impurity or Gini
  importance, providing a ranking of the proxies. The top 100 variables
  from this ranking are manually selected for further modeling
  \citep{schneeweiss2017variable}.
\item
  \textbf{XGBoost}: XGBoost is a gradient boosting algorithm used to
  optimize machine learning models \citep{chen2016xgboost}. It builds
  decision trees that make splits based on maximum impurity reduction,
  and it assigns an importance score to each proxy variable by
  calculating the mean decrease in impurity
  \citep{xiao2024interpretable}.
\item
  \textbf{Stepwise}: Stepwise selection is a progressive feature
  selection method that can proceed in two directions---forward or
  backward---based on the maximum adjusted R-squared. We have
  implemented two versions: (a) Forward selection (FS) starts with an
  initial model (e.g., including all investigator-selected features) and
  adds proxies to the model one at a time. (b) Backward elimination (BE)
  starts with a full model (e.g., all investigator-selected features and
  all proxy variables) and removes features one at a time based on their
  contribution to the model.
\item
  \textbf{Genetic algorithm}: Genetic algorithm is an evolutionary
  algorithm inspired by the theory of natural selection
  \citep{holland1975adaptation}. It operates by evolving offspring from
  a population of the fittest individuals over several generations,
  evaluating and selecting the best combination of features or variables
  that maximize prediction accuracy.
\end{enumerate}

@@

The results for each method under the different scenarios are summarized
below. See Figures \ref{fig:bias-comparison} and
\ref{fig:Coverage-comparison} for an overview of the performance in
terms of bias and coverage, respectively. All simulation results can be
reviewed interactively through a Shiny web application available at
\url{https://ehsanx.shinyapps.io/hdPS-Alternatives/}, providing a
convenient platform for exploring the performance of each method across
various scenarios.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: Bross-based hdPS exhibited the smallest bias (-0.0001),
  and the kitchen sink model (0.0002) was the second. Genetic algorithm
  shows the highest bias (0.0287), indicating a substantial deviation
  from the true effect. Among the other methods, Hybrid hdPS (0.0016),
  and Elastic Net (0.0036) demonstrated low bias. XGBoost (0.0074) had
  slightly higher bias than Random Forest (0.0034).
\item
  \emph{Coverage}: The coverage for most methods was high, with Hybrid
  hdPS, Forward Selection, Backward Elimination, LASSO, and Elastic Net
  achieving values around 98\%, indicating well-calibrated confidence
  intervals. However, Genetic algorithm had noticeably lower coverage
  (83.8\%), indicating that its confidence intervals might be too narrow
  or biased, potentially missing the true effect. After applying bias
  elimination (as there were substantial bias associated with this
  method), Genetic algorithm's coverage improved to 96\%.
\item
  \emph{MSE}: XGBoost achieved the lowest MSE (0.0006). Genetic
  algorithm maintained the highest MSE (0.0016), reflecting its higher
  bias and variability. The kitchen sink model (0.0009), Bross-based
  hdPS (0.0008), Hybrid hdPS (0.0008), and Elastic Net (0.0009) all had
  relatively similar and moderate MSE values.
\item
  \emph{SE}: XGBoost exhibited the lowest Empirical SE (0.0229),
  indicating high precision in its estimates. The kitchen sink model had
  the highest Empirical SE (0.0305), suggesting greater variability.
  Other methods, including Genetic algorithm (0.0274), Hybrid hdPS
  (0.0278), and Bross-based hdPS (0.0287), showed moderate variability.
  LASSO (0.0299) and Elastic Net (0.0294) had slightly higher
  variability. The Model-based SE followed a similar pattern, with
  XGBoost (0.0268) showing the lowest variability and the kitchen sink
  model (0.0333) the highest, indicating less precision in its
  estimates. When comparing relative error in SE estimation, XGBoost
  performed the worst. See Appendix \S C for further details.
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: The kitchen sink model showed a relatively low bias
  (0.0025), while Genetic algorithm continued to exhibit the highest
  bias (0.0408), indicating a significant deviation from the true
  effect. XGBoost had a bias of 0.0259, which, while still higher than
  other methods, but was lower than Genetic algorithm. Other methods,
  such as Bross-based hdPS (0.0035), Hybrid hdPS (0.0049), and Elastic
  Net (0.0053), demonstrated moderate bias. Random Forest (0.0127) and
  Forward Selection (0.0108) had slightly higher bias but remained
  within an acceptable range.
\item
  \emph{Coverage}: Coverage levels remained high for most methods, with
  XGBoost achieving the highest coverage at 96.2\%, indicating
  well-calibrated confidence intervals despite its higher bias. The
  Genetic algorithm method had lower coverage (92.2\%), suggesting that
  its confidence intervals might be narrower, potentially missing the
  true effect. Other methods such as RF, Forward Selection, Backward
  Elimination, and Hybrid hdPS maintained coverage values around
  94-95\%, suggesting adequate interval calibration. Bias-eliminated
  coverage for Genetic algorithm improved to 94.2\%, but it was still
  slightly lower than other methods (e.g., forward selection).
\item
  \emph{MSE}: Forward selection demonstrated the lowest MSE (0.0030),
  and then Hybrid hdPS and XGBoost. The Genetic algorithm method had a
  higher MSE (0.0043), reflecting its substantial bias and variability.
  The kitchen sink model also had an MSE of 0.0043, similar to Genetic
  algorithm, while other methods such as Bross-based hdPS (0.0035), RF
  (0.0039), and Elastic Net (0.0036) exhibited moderate MSE values,
  indicating reasonable accuracy.
\item
  \emph{SE}: The lowest Empirical SE was observed with XGBoost (0.0507)
  and Genetic algorithm (0.0510), reflecting high precision despite
  their higher bias. The kitchen sink model had the highest Empirical SE
  (0.0656), indicating greater variability. Hybrid hdPS (0.0564),
  Bross-based hdPS (0.0595), and RF (0.0609) showed moderate
  variability. Forward Selection (0.0537) and Backward Elimination
  (0.0576) had lower variability compared to the kitchen sink model. In
  terms of Model-based SE, XGBoost (0.0531) and Genetic algorithm
  (0.0533) continued to show low variability, while the kitchen sink
  model had the highest Model-based SE (0.0623), indicating less
  precision in its estimates. When comparing relative error in SE
  estimation, XGBoost and kitchen sink model performed the worst (in
  other direction).
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: In this scenario, the kitchen sink model exhibited a
  moderate negative bias (-0.0093), similar to the Bross-based hdPS
  method (-0.0088). Genetic algorithm showed a significantly higher bias
  (0.0362), indicating a substantial deviation from the true effect.
  Among other methods, XGBoost demonstrated the lowest bias (-0.0061),
  while methods such as Hybrid hdPS (-0.0082), Forward Selection
  (-0.0070), and Backward Elimination (-0.0070) had slightly higher but
  still moderate biases. Elastic Net and LASSO both had biases of
  -0.0079, reflecting slightly larger deviations compared to XGBoost but
  still within acceptable limits.
\item
  \emph{Coverage}: Most methods achieved good coverage, with XGBoost,
  RF, and Forward Selection each achieving a coverage rate of 95.4\%,
  indicating well-calibrated confidence intervals. The Genetic algorithm
  method, however, had slightly lower coverage (91.8\%), indicating that
  its confidence intervals might be narrower, potentially excluding the
  true effect. Bross-based hdPS and the kitchen sink model had slightly
  lower coverage values of 93.8\% and 93.4\%, respectively. After
  accounting for bias, the bias-eliminated coverage for most methods,
  except Genetic algorithm, remained high, with values ranging from
  98.4\% to 99.0\%, indicating that most methods effectively adjusted
  for bias in their coverage estimates. Genetic algorithm's
  bias-eliminated coverage was lower at 93.4\%, reflecting its higher
  inherent bias.
\item
  \emph{MSE}: XGBoost exhibited the lowest MSE (0.0003). Genetic
  algorithm had the highest MSE (0.0040), reflecting its substantial
  bias and variability. The kitchen sink model (0.0005), Bross-based
  hdPS (0.0005), and other methods such as Hybrid hdPS (0.0004) and
  Elastic Net (0.0005) all had relatively similar MSE values, indicating
  moderate accuracy.
\item
  \emph{SE}: The lowest Empirical SE was observed with XGBoost (0.0152),
  reflecting high precision in its estimates. The Genetic algorithm
  method exhibited the highest Empirical SE (0.0523), indicating greater
  variability and less precision. Methods such as Hybrid hdPS (0.0184),
  Forward Selection (0.0187), and Elastic Net (0.0203) showed moderate
  variability, while Bross-based hdPS (0.0206) and the kitchen sink
  model (0.0212) had slightly higher variability. In terms of
  Model-based SE, XGBoost (0.0179) again showed the lowest variability,
  consistent with its low Empirical SE, indicating that it provided the
  most stable estimates. The kitchen sink model had a slightly higher
  Model-based SE (0.0219), indicating less precision in its estimates.
  When comparing relative error in SE estimation, XGBoost performed the
  worst.
\end{enumerate}

@@

See Figure \ref{fig:comparison-plot-updated} for the results from
analyzing the NHANES (2013-2018) dataset. The methods are arranged
according to the number of selected proxy variables. Among all variable
selection algorithms, Random Forest and XGBoost demonstrate the highest
ORs, with values of 1.59 and 1.56, respectively. The ORs for the
remaining methods cluster around 1.52. Additionally, with the exception
of RF and the Bross formula hdPS, a general pattern emerges where
methods selecting a larger number of proxy variables yield lower ORs.
For RD, RF and XGBoost also exhibit the highest value 0.08, while the
remaining methods converge around 0.077. The trend observed in the OR
results appears to persist in the RD analysis.

@@

{\centering \includegraphics[width=1\linewidth,]{figures/ORRD}
